# Required software
* [R](http://cran.r-project.org) is used as the main programming language. Please install at least version: R-4.2.0.
* [RStudio](http://www.rstudio.com/products/rstudio/download/) (optional) is a popular graphic user interface (GUI).

# Installation of the `nfiidd` package

To install the packages needed in the course, including the `nfiidd` package that contains data files used, you can use the `pak` package:

```{r install, eval=FALSE}
install.packages("pak")
pak::pak("nfidd/nfidd", dependencies = "all", upgrade = TRUE)
```

Then you can check that the installation completed successfully by loading the package into your **R** session:

```{r load, eval=FALSE}
library("nfidd")
```

# Installing `cmdstan`

The course relies on running stan through the `cmdstanr` **R** package, which itself uses the `cmdstan` software.
This requires a separate installation step:

```{r cmdstan_install, eval = FALSE}
cmdstanr::install_cmdstan()
```

::: callout-note
This may take a few minutes.
Also you're likely to see lots of warnings and other messages printed to your screen - don't worry, this is normal and doesn't mean there is a problem.
:::

If there are any problems with this, you can try (on Windows) to fix them using
```{r cmdstan_toolchain, eval = FALSE}
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
```

You can test that you have a working `cmdstanr` setup using

```{r cmdstan_test}
cmdstanr::cmdstan_version()
```

For more details, and for links to resources in case something goes wrong, see the [Getting Started with CmdStanr](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) vignette of the package.

# Getting the course materials

You will have to have a local copy of the course materials for the session code to work.
To do so, you can either use git to clone the [repo](https://github.com/nfidd/nfidd), or [download](https://github.com/nfidd/nfidd/archive/refs/heads/main.zip) its current contents.
