Each session in this course uses R code for demonstration.
All the content is self-contained within a software package designed for the course.
To get the most out of this course, you will need to use R and the following instructions to interact with the course material.

# Required software

-   [R](http://cran.r-project.org) is used as the main programming language. Please install at least version: R-4.2.0.
-   [RStudio](http://www.rstudio.com/products/rstudio/download/) is a popular graphic user interface (GUI). Its Visual Editor provides the best experience of going through this course. Please make sure you update RStudio to the latest version.

# Accessing the course

To be able to use the code in each session, you will need a local copy of the course material.

-   Directly download the course material:

    ::: callout-tip
    [[**Download**]{.underline}](https://github.com/nfidd/nfidd/archive/refs/heads/main.zip)
    :::

-   Alternatively, if you are familiar with git you can clone the [repo](https://github.com/nfidd/nfidd).

### Interacting with the course

In this course, all content is written using [R Notebooks](https://bookdown.org/yihui/rmarkdown/notebook.html).
This means that we can combine text with code and see the output directly.
The notebooks are then directly reproduced on the course website (for example, this page).

To interact with each session in the course, we recommend opening the RStudio Project file (`nfidd.RProj`) in the `nfidd` folder you have just downloaded.
Then you can choose to:

-   View each session on the website, and copy-paste the code into your own R script.
    -   Tip: if you hover over each code chunk on the website you can use a "Copy" button at the top right corner.
-   Open the R Notebook for each session.
    -   Each notebook is saved in `nfidd/sessions/` as a `.qmd` file.
    -   Execute code with the single green "play" button at the top-right corner of each code chunk ("Run current chunk"). You can also execute code line-by-line using `Ctrl/Cmd + Enter`.
    -   We suggest using "Visual" view for a better experience (top-left of the RStudio pane).

Before you get started with the course, you will first need to install the following software.

## Installation of the `nfiidd` package

To install the packages needed in the course, including the `nfiidd` package that contains data files used, you can use the `pak` package:

```{r install, eval=FALSE}
install.packages("pak")
pak::pak("nfidd/nfidd", dependencies = "all", upgrade = TRUE)
```

Then you can check that the installation completed successfully by loading the package into your **R** session:

```{r load, eval=FALSE}
library("nfidd")
```

## Installing `cmdstan`

The course relies on running stan through the `cmdstanr` **R** package, which itself uses the `cmdstan` software.
This requires a separate installation step:

```{r cmdstan_install, eval = FALSE}
cmdstanr::install_cmdstan()
```

::: callout-note
This may take a few minutes.
Also you're likely to see lots of warnings and other messages printed to your screen - don't worry, this is normal and doesn't mean there is a problem.
:::

If there are any problems with this, you can try (on Windows) to fix them using

```{r cmdstan_toolchain, eval = FALSE}
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
```

You can test that you have a working `cmdstanr` setup using

```{r cmdstan_test}
cmdstanr::cmdstan_version()
```

For more details, and for links to resources in case something goes wrong, see the [Getting Started with CmdStanr](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) vignette of the package.
