---
title: "Forecasting concepts"
order: 7
bibliography: ../nfidd.bib
---

# Introduction

Epidemiological forecasts are statements about what might happen to population disease burden in the future.
These statements sometimes will be a simple point estimate.
Other times, they might be probabilistic, meaning they are making an explicit statistical statement about the probabilities of different future outcomes.
In this session we will make some simple forecasts using a decades-old statistical modeling approach, Auto-Regressive Integrated Moving Average (ARIMA), which has also shown itself to be a very capable framework for making predictions of infectious diseases.
We will see how we can visualise and interpret such forecasts.

## Slides

-   [Introduction to forecasting](slides/introduction-to-forecasting)

## Objectives

The aim of this session is to introduce the concept of forecasting and forecast visualisation using a simple model.

:::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/forecast-concepts.qmd`.

## Libraries used

In this session we will use the `fpp3` package to load models and methods from the [Forecasting Principles and Practice (v3) textbook](https://otexts.com/fpp3/) (including the `fable` package for simple forecasting models), the `dplyr` package for data wrangling, the `ggplot2` library for plotting and the `epidatr` package for accessing and downloading versions of epidemiological surveillance data from the [Delphi EpiData API](https://cmu-delphi.github.io/delphi-epidata/).

```{r libraries, message = FALSE}
library("fpp3")
library("dplyr")
library("ggplot2")
library("epidatr")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(17)
```
::::

# What is forecasting?

Forecasting is the process of making predictions about the future based on past and present data.
In the context of infectious disease epidemiology, forecasting is usually the process of predicting the future course of some metric of infectious disease incidence or prevalence based on data available at the present time.
In this session we focus on forecasting one particular outcome for influenza surveillance but forecasts can also be made for other quantities of interest such as the number of infections, the reproduction number, or the number of deaths.
Epidemiological forecasting is closely related to nowcasting and, when using mechanistic approaches, estimation of the reproduction number.

In previous sessions, the focus was on a setting where only limited data from a single outbreak was available.
Starting with this session, we will focus on a setting where you have multiple prior "seasons" of data to model.
This introduces new and different opportunities and complexities for modeling.

# Getting acquainted with ILI

Influenza-like illness (ILI) is a commonly used epidemiological surveillance measure for tracking the influenza season.
An individual is classified as having ILI if they have a fever AND an additional "flu-like" symptom, such as a cough, headache, sore throat, etc...
In the US, ILI data has been collected in a more or less consistent way for over 20 years at the national and state level.
These data, including versions available at specific times, are available through the [Delphi EpiData API](https://cmu-delphi.github.io/delphi-epidata/). Specifically, the ILI signal that is published by the CDC measures the percentage of all outpatient doctor's office visits due to ILI in a given epidemiological week.

The following code downloads ILI data for the USA for all weeks starting with the 2023/2024 season through the start of the 2027/2018 season.
Specifically, we use the weighted ILI variable (`wili`) which uses population-based weights estimate to try to generate a representative estimate of ILI for a specific population.
We filter to only include columns that we need and convert the resulting object into a `tsibble` which is [a special kind of time-series data frame](https://tsibble.tidyverts.org/).

```{r load_data}
flu_data <- epidatr::pub_fluview(regions = "nat", 
                                 epiweeks = epirange(200335, 201735)) |> 
  select(region, epiweek, wili) |> 
  as_tsibble(index = epiweek, key = region)
```

::: callout-tip
## What is an epiweek?

Measuring time is complicated.
Epiweeks are a common unit of time in epidemological surveillance data.
Also referred to as [MMWR weeks](https://ibis.doh.nm.gov/resource/MMWRWeekCalendar.html), an epiweek starts on a Sunday and ends on a Saturday.
Most years have 52 epiweeks.
Some have 53.
Week 1 of a specific year is the first week of the calendar year that has at least four days in it.
:::


## Look at the data

Here are the first 10 rows of the dataset.
```{r look-at-data}
flu_data
```

And here is a simple `ggplot`-style figure of the data.
```{r data-first-plot}
ggplot(flu_data) +
  geom_path(aes(x=epiweek, y=wili))
```

But note that one of the advantages of using `tsibbles` is that there are some default time-series plots that can be made:
```{r first-autoplot}
autoplot(flu_data, .vars = wili) +
  labs(title = "National-level ILI data from the US CDC",
       y = "weighted ILI (% of visits)")
```

## Take 5 minutes

Look at the plot of data carefully. Write down at least three observations about the data. Share notes with your neighbor. What features of this time-series make it easier or harder to predict?


# Fitting a simple forecasting model

Before we can forecast we need some data to fit the model to.
In order to assess the quality of the forecasts we will also need some future (or hold-out) data that we can compare the forecasts to.
We will use the same simulated data as in the renewal and nowcasting sessions.
We will try to make a forecast on day 71 (assuming we don't know what the data looks like after that day) as in the nowcasting session.

```{r, load-simulated-onset}
gen_time_pmf <- make_gen_time_pmf()
ip_pmf <- make_ip_pmf()
onset_df <- simulate_onsets(
  make_daily_infections(infection_times), gen_time_pmf, ip_pmf
)
tail(onset_df)
## create hold out dataset
cutoff <- 71
filtered_onset_df <- onset_df |>
  filter(day <= cutoff)
```

## Fitting the model and forecast for 28 days into the future

We can now fit the model to the data and then make a forecast.
This should look very similar to the code we used in the renewal session but with the addition of a non-zero `h` in the data list.

```{r fit_model, results = 'hide', message = FALSE}
horizon <- 28

data <- list(
  n = nrow(filtered_onset_df),
  I0 = 1,
  obs = filtered_onset_df$onsets,
  gen_time_max = length(gen_time_pmf),
  gen_time_pmf = gen_time_pmf,
  ip_max = length(ip_pmf) - 1,
  ip_pmf = ip_pmf,
  h = horizon # Here we set the number of days to forecast into the future
)
rw_forecast <- nfidd_sample(
  mod, data = data, adapt_delta = 0.95,
  init = \() list(init_R = 1, rw_sd = 0.01)
)
```

```{r rw_forecast_summary}
rw_forecast
```

::: callout-note
Because this model can struggle to fit to the data, we have increased the value of `adapt_delta` from its default value of 0.8.
This is a tuning parameter that affects the step size of the sampler in exploring the posterior (higher `adapt_delta` leading to smaller step sizes meaning posterior exploration is slower but more careful).
:::

# Visualising the forecast

We can now visualise the forecast.
We will first extract the forecast and then plot the forecasted number of symptom onsets alongside the observed number of symptom onsets before the forecast was made.

```{r extract-forecast}
forecast <- rw_forecast |>
  gather_draws(forecast[day]) |>
  ungroup() |> 
  mutate(day = day + cutoff)
```

```{r plot_forecast}
forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(data = filtered_onset_df, aes(x = day, y = onsets), color = "black") +
  labs(
    title = "Symptom onsets",
    subtitle = "Forecast (trajectories) and observed (points)"
  )
```

In this plot, the dots show the data and the lines are forecast trajectories that the model deems plausible and consistent with the data so far.

::: callout-tip
## Take 5 minutes

What do you think of this forecast?
Does it match your intuition of how this outbreak will continue?
Is there another way you could visualise the forecast that might be more informative?
:::

::: {.callout-note collapse="true"}
## Solution

-   The forecast mainly predicts things to continue growing as they have been. However, some of the trajectories are going down, indicating that with some probabilities the outbreak might end.
-   Based purely on the data and without knowing anything else about the disease and setting, it is hard to come up with an alternative. The model seems sensible given the available data. In particular, uncertainty increases with increasing distance to the data, which is a sign of a good forecasting model.
-   Instead of visualising plausible trajectories we could visualise a cone with a central forecast and uncertainty around it. We will look at this in the next session as an alternative.
:::

If we want to know how well a model is doing at forecasting, we can later compare it do the data of what really happened.
If we do this, we get the following plot.

```{r plot_future_forecast}
future_onsets <- onset_df |>
  filter(day <= cutoff + horizon)

forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(data = future_onsets, aes(x = day, y = onsets), color = "black") +
  labs(
    title = "Symptom onsets",
    subtitle = "Forecast (trajectories) and observed (points)"
  )
```

::: callout-tip
## Take 5 minutes

What do you think now of this forecast?
Did the model do a good job?
Again, is there another way you could visualise the forecast that might be more informative?
:::

::: {.callout-note collapse="true"}
## Solution

-   On the face of it the forecast looks very poor with some very high predictions compared to the data.
-   Based on this visualisation it is hard to tell if the model is doing a good job but it seems like it is not.
-   As outbreaks are generally considered to be exponential processes it might be more informative to plot the forecast on the log scale.

```{r plot_forecast_log}
forecast |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(alpha = 0.1, aes(y = .value, group = .draw)) +
  geom_point(data = future_onsets, aes(x = day, y = onsets), color = "black") +
  scale_y_log10() +
  labs(title = "Symptom onsets, log scale", subtitle = "Forecast and observed")
```

This should be a lot more informative.
We see that for longer forecast horizons the model is not doing a great job of capturing the reduction in symptom onsets.
However, we can now see that the model seems to be producing very reasonable forecasts for the first week or so of the forecast.
This is a common pattern in forecasting where a model is good at capturing the short term dynamics but struggles with the longer term dynamics.
:::

## Visualising the forecast of the reproduction number

As our forecasting model is based on the reproduction number, we can also visualise the forecast of the reproduction number.
This can be helpful for understanding why our forecasts of symptom onsets look the way they do and for understanding the uncertainty in the forecasts.
We can also compare this to the "true" reproduction number, estimated once all relevant data is available.
To do this, we will fit the model again but with a later cutoff.
Then we can compare the reproduction numbers produced as *forecasts* at the earlier time, with *estimates* at the later time that used more of the data.

```{r fit_model_long, results = 'hide', message = FALSE}
long_onset_df <- onset_df |>
  filter(day <= cutoff + horizon)

long_data <- list(
  n = nrow(long_onset_df),
  I0 = 1,
  obs = long_onset_df$onsets,
  gen_time_max = length(gen_time_pmf),
  gen_time_pmf = gen_time_pmf,
  ip_max = length(ip_pmf) - 1,
  ip_pmf = ip_pmf,
  h = 0
)


rw_long <- nfidd_sample(mod,
  data = long_data,
  adapt_delta = 0.95,
  init = \() list(init_R = 1, rw_sd = 0.01)
)
```

We first need to extract the forecast and estimated reproduction numbers.

```{r extract-Rt}
forecast_r <- rw_forecast |>
  gather_draws(R[day]) |>
  ungroup() |> 
  mutate(type = "forecast")

long_r <- rw_long |>
  gather_draws(R[day]) |>
  ungroup() |>
  mutate(type = "estimate")
```

We can now plot the forecast and estimated reproduction numbers.

```{r plot-Rt}
forecast_r |>
  bind_rows(long_r) |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line(
    aes(y = .value, group = interaction(.draw, type), color = type),
    alpha = 0.1
  ) +
  labs(
    title = "Estimated R",
    subtitle = "Estimated over whole time series (red), and forecast (blue)"
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1)))
```

::: callout-note
The horizontal dashed line at 1 represents the threshold for epidemic growth.
If the reproduction number is above 1 then the epidemic is growing, if it is below 1 then the epidemic is shrinking.
The vertical dashed line represents the point at which we started forecasting.
:::

::: callout-tip
## Take 5 minutes

Can you use this plot to explain why the forecast of onsets looks the way it does?
:::

::: {.callout-note collapse="true"}
## Solution

-   When both models are being fit to data (i.e before the vertical dashed line) the forecast and estimated reproduction numbers are very similar.
-   For short-term forecasts $R_t$ estimates continue to be fairly similar.
-   However, the estimates have a consistent downwards trend which is not captured by the forecast (which looks like it has a constant mean value with increasing uncertainty).
-   This explains the divergence between the forecast and the data as the horizon increases.
-   It looks like only a relatively small number of forecast $R_t$ trajectories grow to be very large but these are enough to visually dominate the forecast of onsets on the natural scale.
-   The performance we are seeing here makes sense given that random walks are defined to have a constant mean and increasing variance.
:::

We managed to learn quite a lot about our model's forecasting limitations just looking at a single forecast using visualisations.
However, what if we wanted to quantify how well the model is doing?
In order to do that we need to look at multiple forecasts from the model.

## Visualising multiple forecasts from a single model

As for a single forecast, our first step is to visualise the forecasts as this can give us a good idea of how well the model is doing without having to calculate any metrics.

```{r plot-all-forecasts}
rw_forecasts |>
  filter(.draw %in% sample(.draw, 100)) |>
  ggplot(aes(x = day)) +
  geom_line(
    aes(y = .value, group = interaction(.draw, origin_day), col = origin_day),
    alpha = 0.1
  ) +
  geom_point(
    data = onset_df |>
      filter(day >= 21),
    aes(x = day, y = onsets), color = "black") +
  scale_color_binned(type = "viridis") +
  labs(title = "Weekly forecasts of symptom onsets over an outbreak",
       col = "Forecast start day")
```

As for the single forecast it may be helpful to also plot the forecast on the log scale.

```{r plot-all-forecasts-log}
rw_forecasts |>
  filter(.draw %in% sample(.draw, 100)) |> 
  ggplot(aes(x = day)) +
  geom_line(
    aes(y = .value, group = interaction(.draw, origin_day), col = origin_day),
    alpha = 0.1
  ) +
  geom_point(data = onset_df, aes(x = day, y = onsets), color = "black") +
  scale_y_log10() +
  scale_color_binned(type = "viridis") +
  labs(title = "Weekly symptom onset forecasts: log scale",
       col = "Forecast start day")
```

::: callout-tip
## Take 5 minutes

What do you think of these forecasts?
Are they any good?
How well do they capture changes in trend?
Does the uncertainty seem reasonable?
Do they seem to under or over predict consistently?
Would you visualise the forecast in a different way?
:::

::: {.callout-note collapse="true"}
## Solution

What do you think of these forecasts?

-   We think these forecasts are a reasonable place to start but there is definitely room for improvement.

Are they any good?

-   They seem to do a reasonable job of capturing the short term dynamics but struggle with the longer term dynamics.

How well do they capture changes in trend?

-   There is little evidence of the model capturing the reduction in onsets before it begins to show in the data.

Does the uncertainty seem reasonable?

-   On the natural scale it looks like the model often over predicts. Things seem more balanced on the log scale but the model still seems to be overly uncertain.

Do they seem to under or over predict consistently?

-   It looks like the model is consistently over predicting on the natural scale but this is less clear on the log scale.
:::

# Going further

## Challenge

-   What other ways could we visualise the forecasts - especially if we had forecasts from multiple models and locations, or both?
-   Based on these visualisations how could we improve the model?

## Methods in practice

-   @heldProbabilisticForecastingInfectious2017 makes a compelling argument for the use of probabilistic forecasts and evaluates spatial forecasts based on routine surveillance data.
-   @hadleyVisualPreferencesCommunicating2025 describes the visualisation of modelling results for national COVID-19 policy and decision makers, with recommendations on what aspects of visuals, graphs, and plots policymakers found to be most helpful in COVID-19 response work.

# Wrap up

-   Review what you've learned in this session with the [learning objectives](../reference/learning_objectives)
-   Share your [questions and thoughts](../reference/help)

## References

::: {#refs}
:::
