---
title: "Forecasting concepts"
order: 7
---

```{r echo = FALSE}
set.seed(123)
```

# Objectives

The aim of this session is to introduce the concept of forecasting, using a simple model, and forecasting evaluation.

# Libraries used

In this session we will use the `nfidd` package to load the data set of infection times, the `dplyr` and `tidyr` packages for data wrangling, `ggplot2` library for plotting, the `here` library to find the stan model, and the `cmdstanr` library for using stan.
We will also use the `tidybayes` package for extracting results of the inference.

```{r libraries, message = FALSE}
library("nfidd")
library("dplyr")
library("tidyr")
library("ggplot2")
library("here")
library("cmdstanr")
library("tidybayes")
```

::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
It needs to be run inside the course repository so that the `here()` commands below find the stan model files.
:::

# Going further

# Wrap up
