---
title: "Local hub playground"
order: 13
bibliography: ../nfidd.bib
---

```{r echo=FALSE}
options(width=100)
```

# Introduction



## Slides

-   [Why use a local hub?](slides/hub-playground)

## Objectives

The aims of this session are 

1. to introduce the use of modeling hubs and hubverse-style tools for local model development and collaborative modeling projects, and
2. to practice building forecasting models using real data.

::: {.callout-note collapse="true"}
# Setup

## Source file

The source file of this session is located at `sessions/hub-playground.qmd`.

## Libraries used

In this session we will use the `dplyr` package for data wrangling, and the `ggplot2` library for plotting.

Additionally, we will use some [hubverse](https://hubverse.io/) packages such as `hubEvals`, `hubUtils`, `hubEnsembles` packages for building ensembles.


```{r libraries, message = FALSE}
library("dplyr")
library("ggplot2")
library("hubUtils")
library("hubEvals")
library("hubEnsembles")
theme_set(theme_bw())
```

::: callout-tip
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(406) # for Ted Williams
```
:::

# A flu forecasting hub for model dev

## Introduce ILI Forecasting at HHS Region level

## Introduce details of this hub


# Building and working with local forecasts

## Building a simple forecast 

### Fitting and generating foreccasts

All locations, all times


### Local hub evaluation



## Building a more complex model


# Submitting forecasts an online hub

## What is your best local model?

Consider models individually, or ensemble

## Process of submitting to an online hub

### GitHub account

### Drag and drop

### Cloning repo, submitting via PR


# References

::: {#refs}
:::
