---
title: "Probability distributions and parameter estimation"
order: 1
---

# Introduction

Many important characteristics, or parameters, of epidemiological processes are not fully observed - and are therefore uncertain. For example, in this course this might include time delays, reproduction numbers, or case numbers now and in the future. We can specify the shape of uncertainty around a specific parameter using a probability distribution. 

We'll want to correctly specify this uncertainty. We can do this best by combining our own prior understanding with the data that we have available. In this course, we'll use this Bayesian approach to modelling. A useful tool for creating these models is the `stan` probabilistic programming language. 

## Slides

-   [Introduction to statistical concepts used in the course](slides/introduction-to-statistical-concepts)
-   [Introduction to stan concepts used in the course](slides/introduction-to-stan)

## Objectives

The aim of this session is to introduce the concept of probability distributions and how to estimate their parameters using Bayesian inference with `stan`.

::: {.callout-note collapse="true"}

# Setup

## Source file

The source file of this session is located at `sessions/R-Stan-and-statistical-concepts.qmd`.

## Libraries used

In this session we will use the `ggplot2` library for plotting, the `here` library to reference files, for example, the stan model, and the `cmdstanr` library for using stan.
We will also use the `bayesplot` and `posterior` packages to investigate the results of the inference conducted with stan.

```{r libraries, message = FALSE}
library("ggplot2")
library("here")
library("cmdstanr")
library("bayesplot")
library("posterior")
```

::: {.callout-tip}
The best way to interact with the material is via the [Visual Editor](https://docs.posit.co/ide/user/ide/guide/documents/visual-editor.html) of RStudio.
If not using the Visual Editor please remember that the code in the session needs to be run inside the course repository so that the `here()` commands below find the stan model files.
:::

## Initialisation

We set a random seed for reproducibility.
Setting this ensures that you should get exactly the same results on your computer as we do.
We also set an option that makes `cmstanr` show line numbers when printing model code.
This is not strictly necessary but will help us talk about the models.

```{r}
set.seed(123)
options(cmdstanr_print_line_numbers = TRUE)
```

:::

# Simulating data from a probability distribution

First let us simulate some data from a probability distribution.
In R, this is usually done using a family of random number generation functions that start with `r`.
For example, to simulate random numbers from a normal distribution you would use the `rnorm()` function.
All these functions have a first argument `n`, the number of random replicates to generate, and then some further arguments that parameterise the probability distribution.

```{r rnorm}
rnorm(n = 10, mean = 0, sd = 1)
```

We will use probability distributions to characterise epidemiological delays.
These are usually strictly positive because e.g. one cannot develop symptoms before becoming infected, and therefore the incubation period cannot be less than zero.
Note that this does not necessarily apply to all distributions, e.g. serial intervals can be negative if person X infects Y but Y develops symptoms first.

Probability distribution that are commonly used in this situation are the gamma or lognormal distributions.
These are fairly similar, with one difference being that the lognormal commonly has a "heavier tail" on the right, i.e., gives more probability to occasional very large values, whereas the gamma distribution has more of a tail on the left, i.e., gives more probability to values lower than the mean.

The gamma distribution is characterised by the `shape` ($\alpha$) and `rate` ($\beta$) parameters, with a mean of $\alpha/\beta$ and variance $\alpha/\beta^2$.
A gamma distribution with mean 5 and variance 2, for example, has $\alpha = 12.5$ and $\beta = 2.5$.
To simulate from such a distribution, we can use the following R code.

```{r gammas}
### simulate gamma with mean 5, variance 2
gammas <- rgamma(100, shape = 12.5, rate = 2.5)
head(gammas)
mean(gammas)
var(gammas)
```

The lognormal distribution is characterised by the `meanlog` ($\mu$) and `sdlog` ($\sigma$) parameters, with a mean of $e^{\mu + 0.5\sigma^2}$ and variance $(e^{\sigma^2} - 1) e^{2\mu + \sigma^2}$.
A lognormal distribution with mean 5 and variance 2, for example, has (after a bit of calculation) $\mu = 1.57$ and $\sigma = 0.28$.

```{r lognormals}
### simulate lognormals with mean 5, variance 2
lognormals <- rlnorm(100, meanlog = 1.57, sdlog = 0.28)
head(lognormals)
mean(lognormals)
var(lognormals)
```

We can now plot the two distributions.

```{r plot_randoms}
df <- rbind(
  data.frame(dist = "lognormal", randoms = lognormals),
  data.frame(dist = "gamma", randoms = gammas)
)
ggplot(df, aes(x = randoms, fill = dist)) +
  geom_density(alpha = 0.5)
```

We have used `geom_density()` to get smooth lines.
Alternatively we could have used, e.g., `geom_histogram()` to plot the raw data.

## Estimating the parameters of probability distributions

We will now use stan to estimate the parameters of the probability distribution.
To do so, we first load in the model.

```{r load_gamma_model}
### load gamma model from the session directory
mod <- cmdstan_model(here("stan", "gamma.stan"))
### show model code
mod
```

::: {.callout-tip}
### Arrays in stan
On line 4 there is a data declaration starting with `array[n]`.
This declares an array of size `n` of the type given aftewards (here: `real`).
Arrays work in a similar way as arrays or vectors in R and its elements be accessed with the bracket operator `[`.
For example, to get the third element of the array `y` you would write `y[3]`.
:::

::: {.callout-tip}
### Take 5 minutes
Familarise yourself with the model above.
Do you understand all the lines?
Which line(s) define the parameter prior distribution(s), which one(s) the likelihood, and which one(s) the data that has to be supplied to the model?
:::

::: {.callout-note collapse="true"}
### Solution
Lines 13 and 14 define the parametric prior distributions (for parameters alpha and beta).
Line 15 defines the likelihood.
Lines 3 and 4 define the data that has to be supplied to the model.
:::

We use the model we have defined in conjunction with the gamma distributed random numbers generated earlier to see if we can recover the parameters of the gamma distribution used.
Once you have familiarised yourself with the model, use the `sample()` function to fit the model.

```{r gamma_inference, results='hide', message=FALSE}
stan_data <- list(
  N = length(gammas),
  y = gammas
)
gamma_fit <- mod$sample(data = stan_data)
```

::: {.callout-note}
### Passing data to the stan model
The `stan_data` object is a list with elements that will is passed to the stan model as the `data` argument to `sample`.
The names and types of the elements need to correspond to the data block in the model (see lines 3 and 4 in the model).
Here, we pass the length of `gammas` as `N` and the vector `gammas` itself as `y`.
:::

::: {.callout-note}
### Stan messages
The`mod$sample` command will produce a lot of messages which we have suppressed above.
This is fine and intended to keep the user informed about any issues as well as general progress with the inference.
This will come in handy later in the course when we fit more complicated models that can take a little while to run.
:::

In order to view a summary of the posterior samples generated, use

```{r gamma_summary}
gamma_fit
```

::: {.callout-caution}
The arguments to `mod$sample()` after the `data` argument are there to remove the amount printed to the screen (and in this document).
You can remove them and you'll get more messages from the stan sampler (which can be very useful for diagnosing and debugging).
:::

You can see that the estimates are broadly consistent with the parameters we specified.
To investigate this further, we will conduct a so-called *posterior predictive check* by comparing random numbers simulated using the estimated parameters to the ones we simulated earlier.

```{r gamma_ppc}
## Extract posterior draws
gamma_posterior <- as_draws_df(gamma_fit$draws())
head(gamma_posterior)

## Generate posterior predictive samples
gamma_ppc <- sapply(seq_along(gammas), function(i) {
  rgamma(n = length(gammas),
         shape = gamma_posterior$alpha[i],
         rate = gamma_posterior$beta[i])
})

## Plot posterior predictive check
ppc_dens_overlay(y = gammas, yrep = gamma_ppc)
```

We can see that the random numbers generated from the posterior samples are distributed relatively evenly around the data (in black), i.e., the samples generated earlier that we fitted to.

## Going further

- For the model above we chose truncated normal priors with a mode at 0 and standard deviation 10. If you change the parameters of the prior distributions, does it affect the results?
- You could try the model included in `lognormal.stan` to estimate parameters of the lognormal distribution.

## Wrap up
