---
title: "Introduction to stan"
author: "Nowcasting and forecasting of infectious disease dynamics"
format:
  revealjs:
    output: slides/introduction-to-stan.html
    footer: "Introduction to stan"
---

```r
set.seed(12345)
```

## What is stan and why do we use it?

- a *Probabilistic Programming Language* for Bayesian inference
  (i.e., a way to write down models)
  
- models are written in a text file (often ending `.stan`) and then loaded into an R/python/etc interface

- once a model is written down, stan can be used to **generate samples** from the **posterior distribution** (using a variety of methods)

## How to write a model in stan {.smaller}

:::: {.columns}

::: {.column width="40%"}
In a stan model file we specify:

- Data<br>(types and names)

- Parameters<br>(types and names)

- Model<br>(prior and likelihood)
:::

::: {.column width="60%"}
<br>

```{stan empty_model, output.var = "empty", eval = FALSE, echo = TRUE}
data {

}

parameters {

}

model {

}
```
:::
::::

## Example: fairness of a coin {.smaller}

:::: {.columns}

::: {.column width="40%"}

Data:

- $N$ coin flips

- $x$ times heads

Parameters

- $\theta$, probability of getting heads; uniform prior in $[0, 1]$

:::

::: {.column width="60%"}
<br>

```{stan coin_model, output.var = "coin", eval = FALSE, echo = TRUE, file = here::here("stan", "coin.stan")}

```
:::
::::

## Using stan from R {.smaller}

There are two packages for using stan from R. We will use the `cmdstanr` package:

```{r load_coin_model, echo = TRUE}
library("cmdstanr")
mod <- cmdstan_model(here::here("stan", "coin.stan"))
mod
```

## Sampling from the posterior {.xmaller}

```{r sample_from_coin_model, echo = TRUE}
data <- list(
  N = 10, ## 10 coin flips
  x = 6 ## 6 times heads
)
mod$sample(data = data)
```

#

[Return to the session](../R-Stan-and-statistical-concepts)
