---
title: "Introduction to the course and the instructors"
author: "Nowcasting and forecasting of infectious disease dynamics"
format:
  revealjs:
    output: slides/introduction-to-the-course-and-the-instructors.html
    footer: "Introduction to the course and the instructors"
---

```{r instructors}
instructors <- c("sbfnk", "seabbs", "kathsherratt")
```

```{r get_contributors}
library("dplyr")
library("gh")
library("jsonlite")
whoarewe <- lapply(sort(instructors), \(instructor) {
  gh(paste0("GET /users/{username}"), username = instructor)[
    c("login", "avatar_url", "html_url", "name")
  ]
})
```

# What is this course?

::: {.incremental}
- delay distributions and how to estimate them (Day 1)
- Rt estimation and the generation interval (Day 1)
- nowcasting (day 2)
- forecasting and evaluation, ensemble methods (day 2)
- applications (day 3)
:::

# Why this course?

::: {.incremental}
- common question in outbreak response
- easy to make mistakes
- no comprehensive training resource
:::

# Who are we?

```{r whoarewe, results='asis'}
cat(":::: {.columns}\n")
for (instructor in seq_along(whoarewe)) {
  cat(
    "::: {.column width=\"", floor(100 / length(whoarewe)), "%\"}\n",
    "![](", whoarewe[[instructor]]$avatar_url, ")\n",
    "[", whoarewe[[instructor]]$name, "](",
    whoarewe[[instructor]]$html_url, ")\n",
    ":::\n", sep = ""
  )
}
cat("::::\n")
```

# Who are you?

Introduce yourself with:

- name
- where you travelled from
- why you're here

#

[Return to the session](../introduction-and-course-overview)
