---
title: "Introduction to statistical concepts used in the course"
author: "Nowcasting and forecasting of infectious disease dynamics"
format:
  revealjs:
    output: slides/introduction-to-statistical-concepts.html
    footer: "Introduction to statistical concepts used in the course"
    slide-level: 3
---

```{r echo = FALSE, message = FALSE}
library("ggplot2")
set.seed(123)
```

### Why statistical concepts?

- We'll need to estimate things (delays, reproduction numbers, case numbers now and in the future)

- We'll want to correctly specify uncertainty

- We'll do this using *Bayesian inference*

## Bayesian inference in 15 minutes

picture

### Interlude: probabilities

::: {.callout-tip}
#### [Laplace, 1812](https://archive.org/details/thorieanalytiqu01laplgoog)

*Probability theory is nothing but common sense reduced to calculation.*
:::

### Interlude: probabilities (1/3) {.smaller}

- If $A$ is a random variable, we write
  $$ p(A = a)$$
  for the *probability* that $A$ takes value $a$.
- We often write
  $$ p(A = a) = p(a)$$
- Example: The probability that it rains tomorrow
  $$ p(\mathrm{tomorrow} = \mathrm{rain}) = p(\mathrm{rain})$$
- Normalisation
  $$ \sum_{a} p(a) = 1 $$

### Interlude: probabilities (2/3) {.smaller}

- If $A$ and $B$ are random variables, we write
  $$ p(A = a, B = b) = p(a, b)$$ for the *joint probability* that $A$ takes value $a$ and $B$ takes value $b$
- Example: The probability that it rains today and tomorrow
  $$ p(\mathrm{tomorrow} = \mathrm{rain}, \mathrm{today} = \mathrm{rain}) = p(\mathrm{rain}, \mathrm{rain})$$
- We can obtain a *marginal probability* from joint probabilities by summing
  $$ p(a) = \sum_{b} p(a, b)$$

### Interlude: probabilities (3/3) {.smaller}

- The *conditional probability* of getting outcome $a$ from random variable $A$, given that the outcome of random variable $B$ was $b$, is written as
  $$ p(A = a | B = b) = p(a| b) $$
- Example: the probability that it rains tomorrow given that it is sunny today
  $$ p(\mathrm{tomorrow} = \mathrm{rain} | \mathrm{today} = \mathrm{rain}) = p(\mathrm{rain} | \mathrm{rain})$$
- Conditional probabilities are related to joint probabilities as
  $$ p(a | b) = \frac{p(a, b)}{p(b)}$$
- We can combine conditional probabilities in the *chain rule*
  $$ p(a, b, c) = p(a | b, c) p(b | c) p (c) $$

### Probability distributions (discrete) {.smaller}

```{r create_poisson_plot}
kicks <- seq(0, 6)
prob <- dpois(kicks, lambda = 0.61)
df <- data.frame(kicks = kicks, prob = prob)
pp <- ggplot(df, aes(x = kicks, y = prob)) +
  theme_bw(20) +
  geom_point(size = 3) +
  ylab("Probability") +
  xlab("Number of kicks")
```

- E.g., how many people die of horse kicks if there are 0.61 kicks per year
- Described by the *Poisson* distribution

:::: {.columns}

::: {.column width="50%"}
```{r plot_poisson1}
pp
```
:::

::: {.column width="50%"}
:::

#### Two directions
1. Calculate the probability
2. Randomly sample


::::

### Calculate discrete probability {.smaller}

- E.g., how many people die of horse kicks if there are 0.61 kicks per year
- Described by the *Poisson* distribution

:::: {.columns}

::: {.column width="50%"}
```{r plot_poisson2}
pp
```
:::

::: {.column width="50%"}
What is the probability of 2 deaths in a year?
```{r dpois, echo = TRUE}
  dpois(x = 2, lambda = 0.61)
```
:::

::::

#### Two directions

1. **Calculate the probability**
2. Randomly sample

### Generate a random (Poisson) sample {.smaller}
- E.g., how many people die of horse kicks if there are 0.61 kicks per year
- Described by the *Poisson* distribution

:::: {.columns}

::: {.column width="50%"}
```{r plot_poisson3}
pp
```
:::

::: {.column width="50%"}
Generate one random sample from the probability distribution
```{r rpois, echo = TRUE}
  rpois(n = 1, lambda = 0.61)
```
:::

::::

#### Two directions

1. Calculate the probability
2. **Randomly sample**

### Probability distributions (continuous) {.smaller}
- Extension of probabilities to *continuous* variables
- E.g., the temperature in Stockholm tomorrow

Normalisation:
$$ \int p(a) da = 1 $$

Marginal probabilities:
$$ p(a) = \int_{} p(a, b) db$$

#### Two directions
1. Evaluate the probability (density)
2. Randomly sample

### Calculate probability density {.smaller}
- Extension of probabilities to *continuous* variables
- E.g., the temperature in Stockholm tomorrow

```{r create_normal_plot}
temp_mean <- 23
temp_sd <- 2
temp <- seq(15, 31, by = 0.01)
prob <- dnorm(temp,mean = temp_mean, sd = temp_sd)
df <- data.frame(temp = temp, prob = prob)
pn <- ggplot(df, aes(x = temp, y = prob)) +
  theme_bw(20) +
  geom_line(lwd = 2) +
  ylab("Temperature") +
  xlab("Probability density")
```

:::: {.columns}

::: {.column width="50%"}
```{r plot_normal1}
pn
```
:::

::: {.column width="50%"}
What is the probability density of $30^\circ C$ tomorrow, if the mean temperature on the day is $23^\circ C$ (standard deviation $2^\circ C$) ? A naïve model could be:
```{r dnorm, echo = TRUE}
  dnorm(x = 30,
        mean = 23,
        sd = 2)
```
:::

::::

#### Two directions

1. **Calculate the probability**
2. Randomly sample

### Generate a random (normal) sample {.smaller}

:::: {.columns}

::: {.column width="50%"}
```{r plot_normal2}
pn
```
:::

::: {.column width="50%"}
Generate one random sample from the normal probability distribution with mean 23 and standard deviation 2:
```{r rnorm, echo = TRUE}
  rnorm(n = 1,
        mean = 23,
        sd = 2)
```
:::

::::

#### Two directions

1. Calculate the probability
2. **Randomly sample**

### Bayesian inference

picture

### Bayes' rule
- We treat the parameters of the a $\theta$ as random with *prior probabilities* given by a distribution $p(\theta)$. Confronting the model with data we obtain *posterior probabilities*  $p(\theta | \mathrm{data})$, our target of inference. Applying the rule of conditional probabilities, we can write this as

$$ p(\theta | \textrm{data}) = \frac{p(\textrm{data} | \theta) p(\theta)}{p(\textrm{data})}$$

- $p(\textrm{data} | \theta)$ is the /likelihood/
- $p(\textrm{data})$ is a /normalisation constant/

- In words,
  $$\textrm{(posterior)} \propto \textrm{(normalised likelihood)} \times \textrm{(prior)}$$

### Bayesian inference

picture

### MCMC

- Markov-chain Monte Carlo (MCMC) is a method to generate *samples* from the *posterior distribution*

- [stan](https://mc-stan.org/) is a statistical software that implements MCMC samplers

#

[Return to the session](../R-Stan-and-statistical-concepts)
