---
title: "Using our Stan models"
subtitle: "Using NFIDD's Stan tools for package models and custom code"
order: 11
---

```{r, include = FALSE}
library(nfidd)
```

This guide covers how to use NFIDD's Stan tools for working with both package models and custom Stan code.

## Using Stan Models

### Basic Package Models

Use models included in the NFIDD package:

```{r}
nfidd_stan_models()
```

Create a model using a package model:

```{r, eval = FALSE}
model <- nfidd_cmdstan_model("simple-nowcast")
```

Sample with course defaults (faster):

```{r, eval = FALSE}
fit <- nfidd_sample(model, data = your_data)
```

### Custom Stan Files

Use a custom Stan file:

```{r, eval = FALSE}
model <- nfidd_cmdstan_model(model_file = "path/to/your/model.stan")
```

You still get access to NFIDD Stan functions:

```{r, eval = FALSE}
fit <- nfidd_sample(model, data = your_data)
```

### Custom Include Paths

Set globally using R options:

```{r, eval = FALSE}
options(nfidd.stan_path = "/path/to/your/stan")
model <- nfidd_cmdstan_model("simple-nowcast")
```

Or override per-model:

```{r, eval = FALSE}
model <- nfidd_cmdstan_model(
  model_file = "custom.stan",
  include_paths = c("/custom/path1", "/custom/path2")
)
```

## Working with Stan Functions

### Discover Available Functions

Get all function names from NFIDD:

```{r}
functions <- nfidd_stan_functions()
functions
```

Find which files contain specific functions:

```{r}
files <- nfidd_stan_function_files(functions = c("renewal", "delay_pmf"))
files
```

### Extract Functions for Local Use

Load specific functions as a string:

```{r}
renewal_code <- nfidd_load_stan_functions(
  functions = c("renewal", "delay_pmf")
)
```

Show first few lines:

```{r}
cat(substr(renewal_code, 1, 200), "...")
```

Write functions to a temporary file for demonstration:

```{r}
temp_file <- file.path(tempdir(), "my_functions.stan")
nfidd_load_stan_functions(
  functions = c("renewal", "delay_pmf"),
  write_to_file = TRUE,
  output_file = temp_file,
  wrap_in_block = TRUE
)
```

Verify file was created:

```{r}
cat("File created at:", temp_file)
cat("\nFile exists:", file.exists(temp_file))
```

### Load All Functions

Get all NFIDD functions:

```{r, eval = FALSE}
all_functions <- nfidd_load_stan_functions()
```

Write all functions to file:

```{r, eval = FALSE}
nfidd_load_stan_functions(
  write_to_file = TRUE,
  output_file = "nfidd_functions.stan",
  wrap_in_block = TRUE
)
```

## Writing Package Models Locally

Sometimes you want to copy a package model to modify it locally rather than modifying the package source.

Load a package model:

```{r, eval = FALSE}
model <- nfidd_cmdstan_model("simple-nowcast", compile = FALSE)
```

Get the Stan code from the model:

```{r, eval = FALSE}
stan_code <- model$code()
```

Write it to a local file:

```{r, eval = FALSE}
writeLines(stan_code, "local-simple-nowcast.stan")
```

Create functions directory:

```{r, eval = FALSE}
dir.create("functions", showWarnings = FALSE)
```

Copy all function files individually:

```{r, eval = FALSE}
stan_functions_path <- file.path(nfidd_stan_path(), "functions")
function_files <- list.files(stan_functions_path, pattern = "\\.stan$", full.names = TRUE)

for (file in function_files) {
  file.copy(file, "functions/", overwrite = TRUE)
}
```

Set options to use local functions:

```{r, eval = FALSE}
options(nfidd.stan_path = ".")
```

Now you can modify the local model and it will use local functions:

```{r, eval = FALSE}
modified_model <- nfidd_cmdstan_model(model_file = "local-simple-nowcast.stan")
```

## Practical Workflows

### Building Custom Models with NFIDD Functions

**Extract the functions you need:**

```{r, eval = FALSE}
nfidd_load_stan_functions(
  functions = c("renewal", "delay_pmf", "convolve_with_delay"),
  write_to_file = TRUE,
  output_file = "my_functions.stan"
)
```

**Create your custom model file:**

```{stan, eval = FALSE, output.var = "custom_model"}
#include my_functions.stan

data {
  // Your data block
}

parameters {
  // Your parameters
}

model {
  // Use NFIDD functions like renewal(), delay_pmf(), etc.
}
```

**Compile and use:**

```{r, eval = FALSE}
model <- nfidd_cmdstan_model(
  model_file = "my_custom_model.stan",
  include_paths = "."
)
```

### Development Workflow

**Explore existing models:**

See what models are available:

```{r}
nfidd_stan_models()
```

Look at model locations:

```{r}
nfidd_stan_path()
```

**Understand function dependencies:**

See all available functions:

```{r, eval = FALSE}
nfidd_stan_functions()
```

Find which files contain functions you need:

```{r, eval = FALSE}
nfidd_stan_function_files(functions = c("function_name"))
```

**Build incrementally:**

Start with package model:

```{r, eval = FALSE}
base_model <- nfidd_cmdstan_model("simple-nowcast")
```

Extend with custom functions:

```{r, eval = FALSE}
custom_model <- nfidd_cmdstan_model(
  model_file = "extended_model.stan"
)
```

## File Organisation

For projects using custom Stan code:

```
your_project/
├── models/
│   ├── my_model.stan
│   └── functions/
│       └── my_functions.stan
```

In your R code:

```{r, eval = FALSE}
options(nfidd.stan_path = c("models/functions", nfidd_stan_path()))
model <- nfidd_cmdstan_model(model_file = "models/my_model.stan")
```

This setup gives you access to both your custom functions and all NFIDD functions.